Light versus dark in strong-lens galaxies: Dark matter haloes that are rounder than their stars

### 2015-08-26 shapes/paper.tex
Justin added references on misalignments
Added/modified Figures, Tables, reply referee

\begin{abstract}
We measure the shape and alignment of the stellar and dark matter mass distribution in 11 strong-lens galaxies, finding over the range of several multiples of the half-light radii that the dark matter haloes are rounder than the stellar mass distributions. Averaging over larger radii, the ellipticity, as defined using the ratio of the semi-minor and semi-major axes, increases both in their dark matter and stellar components. While dark matter haloes are never more elliptical than $e_{dm} = 0.3$, their stars can extend to $e_* > 0.3$. Three systems, in particular, have a high stellar ellipticity ($e_* > 0.4$) and correspondingly a high alignment between the luminous and the dark matter distribution. One of these -- {\it B1608+656} -- is a known merging pair; we suggest that the other two ({\it B0712+472} and {\it B2016+112}) may also be recent post-merger systems. Galaxies with high dark matter ellipticity and weak external shear show strong alignment between light and dark; those with strong shear ($\gamma \gtrsim 0.1$) can be highly misaligned. This is reassuring since isolated misaligned galaxies are expected to be unstable.

@@

The non-parametric mass models themselves are constructed using the
\Glass\ framework for modelling multiple-image lenses
\citep{2014MNRAS.445.2181C}.  \Glass\ produces ensembles of mass maps,
each of which reproduces exactly the observed image positions and
image parities; also time delays, if measured.  The lensing data must
be in the form of point images, which could be quasars or simply
point-like features in otherwise extended images, but this was not a
problem for the lenses studied in this work.  The estimated stellar
mass (explained below in Section~\ref{sec:data}) is taken as a lower
limit on the total mass.  The additional mass that needs to be put on
top, in order to reproduce the lensing observables, is interpreted as
dark matter.  (Some of that mass may actually be gas, but very little
is expected in the early-type galaxies used in this work.)  The
criteria of lensing observables and non-negative dark matter by
themselves would leave the mass distribution too under-determined.
Hence, an additional prior is used, requiring the mass distribution to
be centrally concentrated, with the same centre as the galaxy light
(but not necessarily the shape shape).  The precise formulation of the
prior is given in Section~3 of \cite{2014MNRAS.445.2181C}.  Tests of
how well the shape estimator can be recovered from lensing data are
given in Section~5.2 and Figure~8 of that paper.

@@

\section{Results}\label{sec:results}
\begin{figure*}
  \centering
  \includegraphics[width=.8\linewidth]{Figures/wedges_shears.pdf}
  \caption[width=\linewidth]{The reconstructed lenses are displayed according to the reconstructed total shapes of the dark matter halo $e_{dm}$ and the stellar component $e_{*}$. The wedges show the alignment of the semi-major axes of the components. In case of alignment of the semi-major axes (Aligned), the wedges are vertical. In case of alignment of the semi-major with the semi-minor axes (Anti-Aligned), the wedges are horizontal. The opening angle displays the statistical error in the shape estimate of the dark matter halo. Both the errors in the alignment and the ellipticity of the dark matter distribution denote the ranges $68\%$ of all the mass models lie within. The lenses are additionally color-coded in terms of the required external shear in modelling the lens.}
  \label{fig:wedgesall}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=.8\linewidth]{Figures/wedges.pdf}
  \caption[width=\linewidth]{Similar to Figure~\ref{fig:wedgesall}. In each of the subfigures however we only consider pixels within 1, 2, 3, 4, or 5 $R_e$. If the outermost image of a system lies within one of the limiting radii, the system is kept in this panel, but dropped from subsequent panels.}
  \label{fig:wedgesradii}
\end{figure*}

@@

Figure~\ref{fig:wedgesall} shows our results for the shapes (see \S\ref{sec:shapemethod}) and alignments of the dark matter and stellar mass distributions. In this Figure, the full distribution up to one pixel past the last-image radius $R_{L}$ is probed. There are several key points to note. Firstly, the reconstructed dark matter distribution of all lenses considered here is rounder than the stellar distribution. Secondly, most lenses lie in the bottom-left corner: the dark matter distributions have typically low eccentricities, while the stellar distributions can be rather elliptical. Thirdly, 3 out of the 11 systems, {\it0712}, {\it2016}, and {\it1608}, stand out due to their larger stellar ellipticity ($e_* > 0.4$). At the same time, they show a very strong alignment between their stellar and dark matter distributions. One of these lenses ({\it1608}) is an known merger of an early-type galaxy with a probable late-type galaxy \citep{2003ApJ...584..100S}. We suggest that the other two systems, {\it0712} and {\it2016}, may also be post-merger systems, explaining their similarities with {\it1608}.

Fourthly, notice that all dark matter and stellar distributions with weak external shear (see the contour bar that displays the shear $\gamma$) are aligned apart from {\it0047} which is almost spherical ($e_{dm}\sim0.05$). By contrast, systems requiring a larger external shear ({\it0414}, {\it0911}, and {\it1422}) can display a sizeable misalignment between the distributions. The latter two systems are located in dense environments (see Table~\ref{tab:lensproperties}) that are likely responsible for these misalignments. As described furthermore in \S\ref{sec:data}, {\it1422} is a challenging system to model the stellar distribution. {\it0414} has a perturbed along the line of sight contributing to the lensing effect \citep{2011MNRAS.413L..86C}. Besides these three lens systems, one stands out from the rest, however. In {\it1115}, the semi-major axis of the dark matter seems to be aligned with the semi-minor axis of the stellar component; it is the most strongly misaligned (anti-aligned) lens system of all lenses studied here. This may be explained by its external shear ($\gamma \sim 0.1$) emanating from its group environment (see Table \ref{tab:lensproperties}).

@@

\section{Conclusions}\label{sec:conclusions}
We have measured the shape and alignment of the stellar and dark matter distributions in 11 lens systems using a new non-parametric lens tool, \Glass. We focussed on lenses that have either time delay data or stellar mass maps that contribute significantly to the central potential to determine the projected shape of their dark matter haloes \citep{2014MNRAS.445.2181C}. We measured the shape and alignment using the eigenvalues $\lambda_i$ and eigenvectors of the 2D moment of inertia tensor of the stellar and dark matter distributions, defining a {\it shape parameter} $e = 1 - \lambda_{2}/\lambda_{1}$ ($\lambda_{1} \geq \lambda_{2}$; see \S\ref{sec:shapemethod}). We averaged $e$ over the range 1-5$R_e$, where $R_e$ is the effective radius of the light profile.

@@

\begin{itemize}
\item We found that the dark matter haloes are rounder than the light distribution over the range $1R_e < R < 5R_e$. As we average over larger radii, the lenses, except the ones with a large stellar ellipticity, become increasingly elliptical in their stellar distributions. The same trend can furthermore be seen for the dark matter, although the data are also consistent with no change in shape. The dark matter haloes are never more elliptical than $e_{dm} = 0.2$, while their stars can extend to $e_* > 0.3$.

\item Three systems have a high stellar ellipticity ($e_* > 0.45$) and correspondingly high alignment between light and dark. One of these -- {\it1608} -- is a known merging pair; we suggest that the other two ({\it0712} and {\it2016}) may also be recent post-merger systems.

\item Galaxies with high dark matter ellipticity and weak external shear show strong alignment; those with strong shear ($\gamma \gtrsim 0.1$) can be highly misaligned. This is reassuring since isolated misaligned galaxies are expected to be unstable.
\end{itemize}

@@

>>>>

\begin{abstract}
We measure the shape and alignment of the stellar and dark matter mass distribution in 11 strong-lens galaxies, finding over the range of several multiples of the half-light radii that the dark matter haloes are rounder than the stellar mass distributions. Averaging over larger radii, the ellipticity, as defined using the ratio of the semi-minor and semi-major axes, increases both in their dark matter and stellar components. While dark matter haloes are never more elliptical than $s_{dm} = 0.3$, their stars can extend to $s_* > 0.3$. Three systems, in particular, have a high stellar ellipticity ($s_* > 0.4$) and correspondingly a high alignment between the luminous and the dark matter distribution. One of these -- {\it B1608+656} -- is a known merging pair; we suggest that the other two ({\it B0712+472} and {\it B2016+112}) may also be recent post-merger systems. Galaxies with high dark matter ellipticity and weak external shear show strong alignment between light and dark; those with strong shear ($\gamma \gtrsim 0.1$) can be highly misaligned. This is reassuring since isolated misaligned galaxies are expected to be unstable.

@@

The non-parametric mass models themselves are constructed using the \Glass\ framework for modelling multiple-image lenses \citep{2014MNRAS.445.2181C}. \Glass\ produces ensembles of mass maps, each of which reproduces exactly the observed image positions and image parities; also time delays, if measured. The lensing data must be in the form of point images, which could be quasars or simply point-like features in otherwise extended images, but this was not a problem for the lenses studied in this work. The estimated stellar mass (explained below in Section~\ref{sec:data}) is taken as a lower limit on the total mass. The additional mass that needs to be put on top, in order to reproduce the lensing observables, is interpreted as dark matter. \textbf{(Some of that mass may actually be gas, but very little is expected in the early-type galaxies used in this work.)} The criteria of lensing observables and non-negative dark matter by themselves would leave the mass distribution too under-determined. Hence, an additional prior is used, requiring the mass distribution to be centrally concentrated, with the same centre as the galaxy light (but not necessarily the shape shape). The precise formulation of the prior is given in Section~3 of \cite{2014MNRAS.445.2181C}. Tests of how well the shape estimator can be recovered from lensing data are given in Section~5.2 and Figure~8 of that paper.

@@

\section{Results}\label{sec:results}
\begin{figure*}
  \centering
  \includegraphics[width=.8\linewidth]{Figures/kappaplot.pdf}
  \caption[width=\linewidth]{\textbf{CAPTION}}
  \label{fig:kappaplot}
\end{figure*}

\begin{figure*}
  \centering
  \includegraphics[width=.75\linewidth]{Figures/wedges_shears.pdf}
  \caption[width=\linewidth]{The reconstructed lenses are displayed according to the reconstructed total shapes of the dark matter halo $s_{dm}$ and the stellar component $s_{*}$. The wedges show the alignment of the semi-major axes of the components. In case of alignment of the semi-major axes (Aligned), the wedges are vertical. In case of alignment of the semi-major with the semi-minor axes (Anti-Aligned), the wedges are horizontal. The opening angle displays the statistical error in the shape estimate of the dark matter halo. Both the errors in the alignment and the ellipticity of the dark matter distribution denote the ranges $68\%$ of all the mass models lie within. The lenses are additionally color-coded in terms of the required external shear in modelling the lens. \textbf{update}}
  \label{fig:wedgesall}
\end{figure*}

\begin{table*}
  \begin{center}
    \begin{tabular}{l | l l l l l l}
      Lens & $s_{*}$ & $s_{dm}$ & $e_{*}$ & $e_{dm}$ & $\frac{e_{dm}}{e_{*}}$ & $\frac{e_{dm,t}}{e_{*}}$ \\ \hline
      0047 & 0.27 & 0.04$\pm$0.02 & 0.15 & 0.02$\pm$0.01 & 0.13$\pm$0.07 & 0.10$\pm$0.07 \\
      0414 & 0.19 & 0.06$\pm$0.03 & 0.10 & 0.03$\pm$0.01 & 0.30$\pm$0.14 & 0.20$\pm$0.11 \\
      0712 & 0.43 & 0.08$\pm$0.03 & 0.28 & 0.04$\pm$0.01 & 0.15$\pm$0.05 & 0.15$\pm$0.05 \\
      0911 & 0.22 & 0.07$\pm$0.02 & 0.12 & 0.04$\pm$0.01 & 0.29$\pm$0.10 & 0.27$\pm$0.10 \\
      0957 & 0.19 & 0.10$\pm$0.03 & 0.10 & 0.05$\pm$0.01 & 0.49$\pm$0.14 & 0.49$\pm$0.14 \\
      1115 & 0.30 & 0.11$\pm$0.02 & 0.18 & 0.06$\pm$0.01 & 0.32$\pm$0.07 & 0.08$\pm$0.04 \\
      1422 & 0.05 & 0.06$\pm$0.03 & 0.03 & 0.03$\pm$0.01 & 1.03$\pm$0.51 & 0.64$\pm$0.41 \\
      1608 & 0.47 & 0.28$\pm$0.01 & 0.31 & 0.16$\pm$0.01 & 0.53$\pm$0.03 & 0.52$\pm$0.03 \\
      2016 & 0.47 & 0.12$\pm$0.02 & 0.30 & 0.06$\pm$0.01 & 0.21$\pm$0.04 & 0.20$\pm$0.04 \\
      2045 & 0.17 & 0.03$\pm$0.02 & 0.09 & 0.02$\pm$0.01 & 0.19$\pm$0.11 & 0.19$\pm$0.11 \\
      2237 & 0.24 & 0.05$\pm$0.03 & 0.13 & 0.03$\pm$0.01 & 0.21$\pm$0.10 & 0.21$\pm$0.10 \\
    \end{tabular}
    \caption[width=\linewidth]{...}
    \label{tab:ellipratios}
  \end{center}
\end{table*}

\begin{figure*}
  \centering
  \includegraphics[width=.75\linewidth]{Figures/wedges.pdf}
  \caption[width=\linewidth]{Similar to Figure~\ref{fig:wedgesall}. In each of the subfigures however we only consider pixels within 1, 2, 3, 4, or 5 $R_e$. If the outermost image of a system lies within one of the limiting radii, the system is kept in this panel, but dropped from subsequent panels.}
  \label{fig:wedgesradii}
\end{figure*}

@@

Figure~\ref{fig:wedgesall} shows our results for the shapes (see \S\ref{sec:shapemethod}) and alignments of the dark matter and stellar mass distributions. In this Figure, the full distribution up to one pixel past the last-image radius $R_{L}$ is probed. There are several key points to note. Firstly, the reconstructed dark matter distribution of all lenses considered here is rounder than the stellar distribution. Secondly, most lenses lie in the bottom-left corner: the dark matter distributions have typically low eccentricities, while the stellar distributions can be rather elliptical. Thirdly, 3 out of the 11 systems, {\it0712}, {\it2016}, and {\it1608}, stand out due to their larger stellar ellipticity ($s_* > 0.4$). At the same time, they show a very strong alignment between their stellar and dark matter distributions. One of these lenses ({\it1608}) is an known merger of an early-type galaxy with a probable late-type galaxy \citep{2003ApJ...584..100S}. We suggest that the other two systems, {\it0712} and {\it2016}, may also be post-merger systems, explaining their similarities with {\it1608}.

Fourthly, notice that all dark matter and stellar distributions with weak external shear (see the contour bar that displays the shear $\gamma$) are aligned apart from {\it0047} which is almost spherical ($s_{dm}\sim0.05$). By contrast, systems requiring a larger external shear ({\it0414}, {\it0911}, and {\it1422}) can display a sizeable misalignment between the distributions. The latter two systems are located in dense environments (see Table~\ref{tab:lensproperties}) that are likely responsible for these misalignments. As described furthermore in \S\ref{sec:data}, {\it1422} is a challenging system to model the stellar distribution. {\it0414} has a perturbed along the line of sight contributing to the lensing effect \citep{2011MNRAS.413L..86C}. Besides these three lens systems, one stands out from the rest, however. In {\it1115}, the semi-major axis of the dark matter seems to be aligned with the semi-minor axis of the stellar component; it is the most strongly misaligned (anti-aligned) lens system of all lenses studied here. This may be explained by its external shear ($\gamma \sim 0.1$) emanating from its group environment (see Table \ref{tab:lensproperties}).

@@

\section{Conclusions}\label{sec:conclusions}
We have measured the shape and alignment of the stellar and dark matter distributions in 11 lens systems using a new non-parametric lens tool, \Glass. We focussed on lenses that have either time delay data or stellar mass maps that contribute significantly to the central potential to determine the projected shape of their dark matter haloes \citep{2014MNRAS.445.2181C}. We measured the shape and alignment using the eigenvalues $\lambda_i$ and eigenvectors of the 2D moment of inertia tensor of the stellar and dark matter distributions, defining a {\it shape parameter} $s = 1 - \lambda_{2}/\lambda_{1}$ ($\lambda_{1} \geq \lambda_{2}$; see \S\ref{sec:shapemethod}). We averaged $s$ over the range 1-5$R_e$, where $R_e$ is the effective radius of the light profile.

@@

\begin{itemize}
\item We found that the dark matter haloes are rounder than the light distribution over the range $1R_e < R < 5R_e$. As we average over larger radii, the lenses, except the ones with a large stellar ellipticity, become increasingly elliptical in their stellar distributions. The same trend can furthermore be seen for the dark matter, although the data are also consistent with no change in shape. The dark matter haloes are never more elliptical than $s_{dm} = 0.2$, while their stars can extend to $s_* > 0.3$.

\item Three systems have a high stellar ellipticity ($s_* > 0.45$) and correspondingly high alignment between light and dark. One of these -- {\it1608} -- is a known merging pair; we suggest that the other two ({\it0712} and {\it2016}) may also be recent post-merger systems.

\item Galaxies with high dark matter ellipticity and weak external shear show strong alignment; those with strong shear ($\gamma \gtrsim 0.1$) can be highly misaligned. This is reassuring since isolated misaligned galaxies are expected to be unstable \citep[e.g.][]{1979ApJ...233..872H,1988A&A...206..269M,2007ApJ...670.1027A,2015arXiv150203429D}.
\end{itemize}

%Relevant references for misalignments being unstable: 
%- http://adsabs.harvard.edu/abs/1978MNRAS.183..779B : 1978MNRAS.183..779B
%- http://adsabs.harvard.edu/abs/1981MNRAS.196..455B : 1981MNRAS.196..455B
%- http://adsabs.harvard.edu/abs/1979ApJ...233..872H : 1979ApJ...233..872H
%- http://adsabs.harvard.edu/abs/1984ApJ...276..101S : 1984ApJ...276..101S
%- http://adsabs.harvard.edu/abs/1987A%26A...182..202R : 1987A&A...182..202R
%- http://adsabs.harvard.edu/abs/1988A%26A...206..269M : 1988A&A...206..269M
%- http://adsabs.harvard.edu/abs/2007ApJ...670.1027A : 2007ApJ...670.1027A

@@

\section{Shear-Shape degeneracy}\label{sec:shearshapedeg}
\begin{figure*}
  \centering
  \includegraphics[width=.9\linewidth]{Figures/theta_scatter.pdf}
  \caption[width=\linewidth]{\textbf{CAPTION}}
  \label{fig:thetascatter}
\end{figure*}
